%!TEX TS-program = xelatex
%!TEX encoding = UTF-8 Unicode

\documentclass[a4paper]{article}

\usepackage{xltxtra}
\usepackage{amsfonts}
\usepackage{polyglossia}
\usepackage{fancyhdr}
\usepackage{geometry}
\usepackage{dsfont}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{physics}
\usepackage{mathtools}

\geometry{a4paper,left=15mm,right=15mm,top=20mm,bottom=20mm}
\pagestyle{fancy}
\lhead{Devon Morris}
\chead{Differential Geometry}
\rhead{\today}
\cfoot{\thepage}

\setlength{\headheight}{23pt}
\setlength{\parindent}{0.0in}
\setlength{\parskip}{0.0in}

\newtheorem*{prop}{Proposition}
\newtheorem*{defn}{Definition}
\newtheorem*{thm}{Theorem}
\newtheorem*{cor}{Corollary}
\newtheorem*{lem}{Lemma}
\newtheorem*{rem}{Remark}

\DeclarePairedDelimiterX{\inn}[2]{\langle}{\rangle}{#1, #2}

\begin{document}
\section*{Geodesics}%
Let $(M,g)$ be a Riemannian manifold with $\nabla$ the Levi-Civita connection.

\begin{defn}
  A smooth curve $\gamma: [a,b] \rightarrow M$ is a geodesic if $\frac{D}{dt} \left( \dv{\gamma}{t} \right) =0$ at every $t \in I$. In other words $\dv{\gamma}{t}$ is parallel along $\gamma$.
\end{defn}
These are the analog of ``straight-lines'' on a manifold. We can think of this parallel requirement in terms of not allowing the velocity to change. Equivalently we could say these curves have 0 acceleration and the entire trajectory is determined by initial position and velocity. 

\begin{rem}
 By the compatibility of $\nabla$ with $g$ we have 
 \[
   \dv{t} g \left( \dv{\gamma}{t}, \dv{\gamma}{t} \right) = g \left( \frac{D}{dt} \left( \dv{\gamma}{t} \right),  \dv{\gamma}{t} \right) + g \left(\dv{\gamma}{t}, \frac{D}{dt} \left( \dv{\gamma}{t} \right)\right) = 0
 \]
 So the length of our velocity vector $\norm{\dv{\gamma}{t}}^2 = c^2$ is constant. Thus the length of a geodesic $\gamma: [a,b] \rightarrow M$
 \[
   \ell_a^b(\gamma) = \int_a^b \sqrt{ g \left( \dv{\gamma}{t}, \dv{\gamma}{t} \right)}\ dt = c(b-a)
 \]
 So the arclength of a geodesic is proportional to parameter length.
\end{rem}

Let $(U,\varphi) = (U, (x^j))$ be a coordinate chart on $M$ and $\gamma([a,b]) \cap U \neq \varnothing$, Then $\gamma(t) = (\gamma^1(t), \dots, \gamma^n(t))$ and 
\[
  \dv{\gamma}{t} = \dv{\gamma^j}{t} \pdv{x^j}
\]
hence 
\[
  0 = \frac{D}{dt} \left( \dv{\gamma}{t} \right) = \left(\dv[2]{t} \gamma^k + \dv{\gamma^j}{t}\dv{\gamma^i}{t} \Gamma_{ij}^k\right) \pdv{x^k}
\]
for all $k$. To solve $\dv{\gamma^k}{t} = y^k$ and solve the resulting first order system
\[
  \begin{aligned}
    \dv{\gamma^k}{t} &= y^k  \\
    \dv{y^k}{t} &= -\Gamma_{ij}^ky^jy^i
  \end{aligned}
\]
We want to solve for $(\gamma^1, \dots, \gamma^n, y^1, \dots, y^n) = \left( \gamma^1, \dots, \gamma^n, \dv{\gamma^1}{t}, \dots, \dv{\gamma^n}{t} \right)$. Recall that if $(U, \varphi) = (U, (x^j))$ are coordinates on $M$ and $\pi: TM \rightarrow M$ is the projection, then $\varphi$ induces a chart $\bar{\phi}$ on $\pi^{-1}(U)$ by 
\[
  (p,v) \mapsto (x^1(p), \dots, x^n(p), v^1, \dots, v^n)
\]
where $v \in T_pM$ is $v = v^j \pdv{x^j}$. Then for any $(p,v) = \left(\gamma^1(t_0), \dots, \gamma^n(t_0),\dv{\gamma^1}{t}(t_0), \dots, \dv{\gamma^n}{t}(t_0)\right)$, then the system above has a unique solution. In other words, for any $(p,v) \in TM$, there is a unique geodesic $\gamma$ passing through $p$ with $\dv{\gamma}{t} = v$.

\begin{lem}
  There exists a unique vector field $G$ on $TM$, $G \in \Gamma^{\infty}(TTM)$, whose integral curves are all of the form $t \mapsto (\gamma(t), \dv{\gamma}{t}(t))$, where $\gamma$ is a geodesic.
\end{lem}

Recall that $X \in \Gamma^{\infty}(N)$, for all $p \in N$, there exists $U$ containing $p$, and $\delta >0$, an a map $\theta: (-\delta, \delta) \times U \rightarrow N$ such that for all $q \in U$, the curve $\theta_{q}(t) = \theta(t,q)$ with 
\[
  \dv{\theta_q}{t} = X_{\theta_q(t)}
\]
with $\theta(0,q) = q$. $\theta$ is the local flow generated by $X$.

\begin{defn}
  The flow on $TM$ generated by $G$ is called the geodesic flow.
\end{defn}

\begin{lem}
  If the geodesic $\gamma(t,p,v)$ is defined for $t \in (-\delta, \delta)$ and $a > 0$ thenthe geodesic $\gamma(at, p,v)$ is defined on the interval $(-\frac{\delta}{a}, \frac{\delta}{a})$ and $\gamma(a t, p, v) = \gamma(t, p, av)$.
\end{lem}

\begin{proof}
  Let $h : (-\frac{\delta}{a}, \frac{\delta}{a}) \rightarrow M$ be defined by $h(t) = \gamma(at, p, v)$. Then $h(0) = \gamma(0, p,v) = p$, 
  \[
    \dv{h}{t}(0) =  a \dv{\gamma}{t} (at, p, v) = av
  \]
  We should check that it is a geodesic. So locally
  \[
    \frac{D}{dt} \left(\dv{h}{t}\right) = \nabla_{\dv{h}{t}} \dv{h}{t} =  \nabla_{a \dv{\gamma}{t}} \dv{\gamma}{t} = a^2 \nabla_{\dv{\gamma}{t}} \dv{\gamma}{t} = 0
  \]
  since $\gamma$ is a geodesic. Thus $h(t)$ is a geodesic passing through $p$. Thus by uniqueness we must have that
  \[
    h(t) = \gamma(at, p, v) = \gamma(t, p, av)
  \]
\end{proof}

\begin{defn}
  Let $M_0$ be the zero section of the tangent bundle $TM$, 
  \[
    M_0 = \coprod_{p \in M} \left\{ 0_p \right\}
  \]
  $0_p \in T_pM$.
\end{defn}

So using the zero section we can identify the zero section with the manifold itself. We can use this lemma.

\begin{cor}
  There is an open neighborhood $U$ of $M_0$ in $TM$ so that for every $(p,v) \in U$ the geodesic $t \mapsto \gamma(t,p,v)$ is defined for all $t \in (-2,2)$.
\end{cor}

\begin{proof}
  Around each $p \in M_0$, pick a neighborhood $U_p$ on which the geodesicsare defined for all $(-\delta_p, \delta_p)$. We can use the above lemma to extend $(-\delta_p, \delta_p)$ to $(-2, 2)$ at the expense of shrinking down the neighborhood $U_p$. The let $U = \cup_{p \in M} U_p$.
\end{proof}

\begin{defn}
  The map $\exp: U \rightarrow M$ defined by 
  \[
    \exp: (p,v) \mapsto \gamma(1, p, v) =  \gamma(|v|, p, v/|v|)
  \]
  is called the exponential map.
\end{defn}
We'll often restrict $\exp$ to a neighborhood of $0_p$ in some $T_pM$. Let $B_{\epsilon}(0)$ be the ball of radius $\epsilon$ about $0_p \in T_pM$. For $\epsilon$ sufficiently small define 
\[
  \begin{aligned}
    \exp_p: B_{\epsilon}(0) \rightarrow M \\
    (p,v) \mapsto \exp(p,v)
  \end{aligned}
\]

\begin{prop}
  Given $p \in M$, there exists $\epsilon > 0$, such that $\exp_p: B_{\epsilon}(0) \rightarrow M$ is a diffeomorphism from $B_{\epsilon} \subset T_pM$ onto an open subset of $M$. 
\end{prop}

Before we do the proof we state the following remark.

\begin{rem}
  For any finite dimensional vector space $V$ any $v \in V$, there is a canonical identification $ V \cong T_vV$. It is given by $w \in V$ is given by $v + tw$ and we think of $w$ as $[v + tw] = \dv{v}{t} (v + tw)$ 
\end{rem}

\begin{proof}
  We will show that the differential $d (\exp_p)_0$ is an isomorphism
  \[
    d(\exp_p)_0 : T_0T_pM \rightarrow T_pM
  \]
  let $\tilde{w} \in T_0T_pM$, we can identify $\tilde{w}$ with $[tw] = \dv{t}(tw)$ for some $w \in T_pM$. So we have
  \[
    d(\exp_p)_0(\tilde{w}) = \dv{t}\eval{(\exp_0(tw))}_{t=0} = \dv{t} \eval{\left( \gamma(1, p, tw)\right)}_{t=0} = \dv{t} \eval{\left( \gamma(t, p, w) \right)}_{t=0} = w
  \]
  So $d(\exp_p)_0$ is an isormorphism, and $\exp_p$ is a diffeomorphism on a neighborhood of $0 \in T_pM$ by the invertible mapping theorem.
\end{proof}

\begin{defn}
  Let $p \in M$ and let $B_{\epsilon}(0) \subset T_pM$ be an open ball such that $\exp_p$ restricts to a diffeomorphism on some open set $V$ with the closure of the $\overline{B_{\epsilon}(0)} \subset V$. Then, 
  \[
    B_{\epsilon}(p) = \exp_p (B_{\epsilon}(0))
  \]
  is called the normal ball (or geodesic ball) center at $p$ with radius $\epsilon$.
\end{defn}
 
So $\exp_p(v)$ is called a normal neighborhood of $p$. We call $\partial B_{\epsilon}(p)$ is calld normal (or geodesic) sphere. Geodesics in $B_{\epsilon}(p)$ which start at $p$ are radial geodesics.

\begin{defn}
  A piecewise differentiable or $C^{\infty}$ curve is a continuous map $c:[a,b] \rightarrow M$ such that there exists $a= t_0 < t_1 < \dots < t_n = b$ with $\eval{c}_{(t_i, t_{i+1})}$ is smooth for all $i$. The points $c(t_i)$ are called vertices of $c$ and the angle
  \[
    \lim_{t \rightarrow t_i^{-}} \dv{c}{t} \text{ and } \lim_{t \rightarrow t_i^+} \dv{c}{t}
  \]
  is called the vertex angle of $c(t_i)$. 
\end{defn}
This assumes that we have a connection (parallel transport). We want to prove the following.

\begin{lem}
  Let $p \in M$, $U$ a normal neighborhood of $p$ and $B \subset M$ a normal ball centered at $p$. Let $\gamma: [0,1] \rightarrow B$ be a geodesic with $\gamma(0) = p$ if $c:[0,1] \rightarrow M$ is some other piecewise differentiable curve with $c(0) = \gamma(0)$ and $c(1) = \gamma(1)$ then $\ell(c) \geq \ell(\gamma)$ and if equality holds then $\gamma([0,1]) = c([0,1])$.
\end{lem}

\begin{lem}[Gauss' Lemma]
  Let $p \in M$ and let $v \in T_pM$ such that $\exp_p(v)$ is defined. Let $w \in T_pM \cong T_v(T_pM)$. Then 
  \[
    \inn*{d(\exp_p)_v(v)}{d(\exp_p)_v(w)} = \inn*{v}{w}
  \]
  (on left hand we think $v,w \in T_v(T_pM)$ and on right hand $v,w \in T_pM$).
\end{lem}
We can think of this as saying that the radial geodesics are orthogonal to the normal spheres.

\begin{proof}[proof: of previous lemma]
  Suppose first that $c([0,1]) \subset B$. For all $t \in (0,1]$, we can write $c(t) = \exp_p(r(t)v(t))$, where $t \mapsto v(t) \in T_pM$ is a curve in $T_pM$ with $|v(t)| = 1$ for all $t$, and $r: (0,1] \rightarrow \mathds{R}$ is some piecewise $C^{\infty}$ function with $r(t) > 0$ for all $t$. We can assume that $c(t) \neq p$ for $t \in (0,1]$. We write $f(r,t) = \exp_p(r v(t))$, then $c(t) = f(r(t),t)$, then except at a finite number of points we have 
  \[
    \dv{c}{t} = \dv{f(r(t), t))}{t} = \pdv{f}{r}\dv{r}{t} + \pdv{f}{t}
  \]
  where we define
  \[
    \pdv{f}{r} = df \left( \pdv{r} \right) \text{ and }  \pdv{f}{t} = df \left( \pdv{t} \right)
  \]
  $\pdv{r}$ is a radial vector field on $T_pM$ and $\pdv{f}{r} = df \left( \pdv{r} \right) = d\left(\exp_p(r v(t))\right) \left( \pdv{r} \right)$ will be orthogonal to $\pdv{f}{t}$ which is
  \[
    \pdv{f}{t} = df \left( \dv{v}{t} \right)
  \]
  since $|v(t)| = 1$, by gauss's lemma. So
  \[
    \inn*{\pdv{f}{r}}{\pdv{f}{t}} = 0
  \]
  also $|\pdv{f}{r}| = 1$ since 
  \[
    \pdv{f}{r} = \pdv{r} \left( \exp_p \left( r(t) v(t) \right) \right) = \pdv{r} \left( \gamma(1, p, r(t)v(t) \right) = \pdv{r} \left( \gamma(r(t), p, v(t)) \right)  = v(t)
  \]
  and $|v(t)| = 1$. So we have that
  \[
    \left| \dv{c}{t} \right|^2 = \left| \pdv{f}{r} \right| \left| \dv{r}{t} \right| + \left| \dv{r}{t} \right| + \left| \pdv{f}{t} \right|^2 \geq \left| \dv{r}{t} \right|^2
  \]
  So
  \[
    \int_{\epsilon}^1 \left| \dv{c}{t} \right| dt \geq \int_{\epsilon}^1 \left| \dv{r}{t} \right| dt \geq \int_{\epsilon}^1 \dv{r}{t} dt = r(1) - r(\epsilon)
  \]
  So we have that
  \[
    r(1) = \ell(\gamma)
  \]
  Therefore we have that
  \[
    \ell(c) \geq \ell(\gamma)
  \]

  \begin{defn}
    A subset $S \subset M$ is strongly convex if given any two $p_1,p_2 \in \bar{S}$, there exists a unique length minimizing geodesic joining $p_1, p_2$ whose interior lies in $S$. 
  \end{defn}
  \begin{prop}
    every $p \in M$ has strongly convex neighborhood.
  \end{prop}
\end{proof}

We want to show that length minimizing curves are geodesics. 

\begin{lem}
  For every point, $p \in M$, there exists a neighborhood $W$ containing $p$ and some $\delta > 0$, such that for all $q \in W$, $\exp_q$ is a diffeomorphism on $B_{\delta}(0)$, and $W \subset \exp_q(B_{\delta}(0))$. Therefore, $W$ is a normal neighborhood of each of its points.
\end{lem}

\begin{proof}
  Let $U \subset TM$ be as in the previous lemma (where we get the existence of the geodesic up to time 2). Define 
  \[
    \begin{aligned}
      F: &U \rightarrow M \times M \\
         &(q,v) \mapsto (q, \exp_q(v))
    \end{aligned}
  \]
  By looking at $F$ in some local coordinates,
  \[
    \eval{dF}_{(p,0)} = 
    \begin{bmatrix}
      I & 0 \\
      *  & I
    \end{bmatrix}
  \]
  Since $d(\exp_p)_0 = \text{Id}$ as shown in a previous lemma. Thus, there exists a neighborhood $\tilde{U} \subset U$ of the point $(p,0)$ and $U',W'$ in $M \times M$ such that $\eval{F}_{U'}$ is a diffeomorphism. We can assume $U' = \left\{ (q,v): q \in V' \text{ and } v \in T_qM, |v| < \delta \right\}$ for some open $V' \subset M$ and some $\delta > 0$. Choose a neighborhood $W \subset M$ with $p \in W$, such that  $W \times W$ is contained in $W'$. If $q \in W$ and $B_{\delta}(0) \subset T_qM$, then $q \in V'$ and $\left\{ q \right\} \times B_{\delta}(0) \subset U'$ and since $F$ is a diffeomorphism on $U'$ we have $\left\{ q \right\} \times W \subset F(\left\{ q \right\} \times B_{\delta}(0)).$ 
\end{proof}

Note: This implies that for all $q_1, q_2 \in W$ there exists a unique geodesic $\gamma$ of length $\leq \delta$, joining $q_1$ to $q_2$. Also since the $F^{-1}(q_1, q_2) = (q_1, v)$, the starting velocity vector $v$ of $\gamma$ depends smoothly on $q_1, q_2$. So $W$ is called a totally normal neighborhood of $p \in M$.

\begin{cor}
  If a piecewise diff curve $\gamma: [a,b] \rightarrow M$ with parameter proportional to arclength has $\ell(\gamma) \leq \ell(c)$ for any other piecewise differentiable curve $c$.with $c(a) = \gamma(a)$, $c(b) = \gamma(b)$ then $\gamma$ is a geodesic.
\end{cor}

\section*{Riemannian Manifolds as Metric Spaces}%
In this section assume that $M$ is connected.

\begin{defn}
 The Riemannian distance function on $d:M\times M \rightarrow M$ is  
 \[
   d(p,q) = \inf \left\{ \ell(c): c\text{ is a piecewise differentiable curve joining } p \text{ and } q \right\}
 \]
\end{defn}

\begin{thm}
  $d: M \times M \rightarrow \mathds{R}$ defines a metric on $M$ (in the sense that $(M,d)$ is a metric space) and the metric topology on $M$ is the same as the given topology.
\end{thm}

\begin{proof}
  The first part of the proof was done in homework. Since geodesics are length minimizing the balls of $d(\cdot, \cdot)$ are geodesic ball, which are open in $M$ since they are the images of open sets under a diffeomorphism $\exp_p$.
\end{proof}

\begin{cor}
  $d: M \times M \rightarrow \mathds{R}$ is continuous. 
\end{cor}

\begin{defn}
  A Riemannnian Manifold id geodesically complete  if for all $p \in M$ the exponential map $\exp_p: T_pM \rightarrow M$ is defined for all $v \in T_pM$ i.e. the geodesics starting at the point $p$ exist for all time.
\end{defn}

\begin{defn}
  A Riemannian manifold $M$ is extendable if there is a connected Riemannian manifold $M'$ such that $M$ is isometric to a proper open subset of $M'$. If not, $M$ is non-extendable.
\end{defn}

\begin{prop}
  If $M$ is geodesically complete then $M$ is non-extendable.
\end{prop}

\begin{proof}
  Suppose that $M \subset M'$ and is open. Pick some $p \in \partial M \subset M'$ and choose a geodesic ball $B \subset M'$ about $p$. Let $q \in B \cap  M$, let $\gamma$ be a geodesic in $M'$ with $\gamma(0) = p$ and $\gamma(1) = q$. Then $\tilde{\gamma}(t) = \gamma(1-t)$ is a geodesic with $\tilde{\gamma}(0) = q$ and $\tilde{\gamma}(1) = p$ which is not in $M$. Thus is $M$ is not geodesically complete, by uniqueness of geodesics.
\end{proof}

Note that the class of nonextendable manifolds is larger than the class of geodesically complete manifolds. 

\begin{thm}[Hopf-Rinow]
 Let $M$ be a Riemannian manifold with $p \in M$. Then the following are equivalent  
 \begin{enumerate}
   \item $\exp_p:T_pM \rightarrow M$ is defined on all $T_pM$ 
   \item The closed bounded subsets of $M$ are compact 
   \item $M$ is complete as a metric space.
   \item $M$ is geodesically complete
   \item There exists a covering of $M$ by nested compact sets, such that if $p_n \notin k_n$ then $\lim_{n \rightarrow \infty} d(p,q_n) = \infty$
 \end{enumerate}
 Each of these implies that 6. For any $q \in M$ there exists a geodesic  $\gamma$ joining $p$ and $q$.
\end{thm}

\begin{proof}
  (1 $\implies$ 6) Let $q \in M$, $d(q,p) = r$. Let $B_{\delta}(p)$ be a geodesic ball with $S = \partial B_{\delta}(p)$. Let $x_0 \in S$ be the point which minimizes $d(q,S) = \left\{ d(q,x); x \in S \right\}$. Then $x_0 = \exp_p(\delta v)$ for some $v \in T_pM$ with $|v|=1$. Let $\gamma(s) = \exp_p(sv)$ be the geodesic from $p$ to $x_0$. We want to show that $\gamma(r) = q$. Let $A = \left\{ s \in \left[ 0,r \right]: d(\gamma(s), q) = r-s \right\}$. First we note that $0 \in A$, so $A \neq \varnothing$. $A$ is closed in $[0,r]$ by continuity. (Here we used 1 in the fact that $\gamma$ is defined on all of $[0,r]$). $A \subset [0,r]$ is compact and has a maximal element, say $s_0$. Now assume that $s_0 < r$. Let $\delta'> 0$  such that $B_{\delta'}(\gamma(s_0))$ is a geodesic ball with $s_0 + \delta' < r$. Let $x_0'$ be the point on $\partial B_{\delta'}(\gamma(s_0))$ which minimizes the distance to $q$. Note that 
\[
  r - s_0 =  d(\gamma(s_0), q) = \delta' + \min_{x\in \partial B_{\delta'}} d(x,q) = \delta' + d(x_0', q)
\]
Then by the triangle inequality we have
\[
  d(p,x_0') \geq d(p,q) - d(x_0',q) = r - (r - s_0 - \delta') = s_0 \delta'
\]
The broken curve joining $p$ to $x_0$ by joining $p$ to $\gamma(s_0)$ along $\gamma$ then joining $\gamma(s_0)$ to $x_0'$ along a geodesic ray has length $s_0 + \delta'$, so it is a minimizing curve and hence it is an unbroken geodesic. It is really just an extension of $\gamma$. So, $\gamma(s_0 + \delta') = x_0$, since $r - s_0 = \delta' + d(x_0',q) = \delta' + d(\gamma(s_0 + \delta'), q)$. Therefore we have that
\[
  d(\gamma(s_0 + \delta'), q) = r - (s_0 + \delta')
\]
Therefore $s_0 + \delta' \in A$, creating a cotradiction. So $s_0 = r$ implying that $\gamma(r) = q$.

(1 $\implies$ 2) Let $A \subset M$ be closed and bounded then $A \subset B$ for some metric ball $B$. By $(6)$, therefore $B$ is in the image of $\exp_p$ and $B \subset \exp_p\left(\overline{B_r(0)}\right)$ for some $r$ but as $\exp_p$ is continuous and the closure of this ball is compact we know that $\exp_p \left( \overline{B_r(0)} \right)$ and since $A$ is a closed subset of a compact set $A$ is compact.
\end{proof}

\end{document}
