%!TEX TS-program = xelatex
%!TEX encoding = UTF-8 Unicode

\documentclass[a4paper]{article}

\usepackage{xltxtra}
\usepackage{amsfonts}
\usepackage{polyglossia}
\usepackage{fancyhdr}
\usepackage{geometry}
\usepackage{dsfont}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{physics}
\usepackage{mathtools}

\geometry{a4paper,left=25mm,right=25mm,top=25mm,bottom=25mm}
\pagestyle{fancy}
\lhead{Devon Morris}
\chead{Differential Geometry}
\rhead{\today}
\cfoot{\thepage}

\setlength{\headheight}{23pt}
\setlength{\parindent}{0.0in}
\setlength{\parskip}{0.0in}

\newtheorem*{prop}{Proposition}
\newtheorem*{defn}{Definition}
\newtheorem*{thm}{Theorem}
\newtheorem*{cor}{Corollary}
\newtheorem*{lem}{Lemma}
\newtheorem*{rem}{Remark}

\DeclarePairedDelimiterX{\inn}[2]{\langle}{\rangle}{#1, #2}

\begin{document}

\section*{Lie Groups}%

Let $G$ be a lie group. Recall that we have the two mappings given by 
\[
  \begin{aligned}
    L_x: &G \rightarrow G \\ 
       &y \mapsto xy \\
    R_x: &G \rightarrow G \\
         &y \mapsto yx
  \end{aligned}
\]
and that There is a tangent space at the identity $T_eG$ which we call the lie algebra and denote it by $\mathcal{G}$.

\section*{Left and Right Invariant Vector Fields}%
Draw the picture of the circle with an invariant vector field.
\begin{defn}
Let $X \in \Gamma^{\infty}(TG)$, we call $X$ left invariant if 
\[
  dL_x(X_y) = X_{xy}
\]
for all $x,y \in G$, similarly $X$ is right invariant if
\[
  dR_x(X_y) = X_{yx}
\]
for all $x,y \in G$. Or for simplicity $dL_xX = X$.
\end{defn}

\begin{rem}
  This gives us a different interpretation of the lie algebra $\mathcal{G}$ given any $v \in \mathcal{G}$, $v$ generates a left (right) invariant vector $X \in \Gamma^{\infty}(TG)$ by
  \[
    X_y = dL_y(v) \quad \text{ or } \quad \left(X_y = dR_y(v) \right)
  \]
  It should be easy to see that this is left invariant.
  \[
    d L_x(X_y) = dL_x(dL_y(v)) = dL_{xy}(v) = X_{xy}
  \]
  Similarly for right invariance. This allows us to interpret vectors in the lie algebra $\mathcal{G}$ as left (right) invariant vector fields on the manifolds.
\end{rem}

\begin{prop}
  The lie bracket of two left (right) invariant vector field $X,Y \in \Gamma^{\infty}(TG)$ is also left (right) invariant
\end{prop}

\begin{proof}
  \[
    \begin{aligned}
      dL_x [X,Y]f &=  [X,Y](f \circ L_x) = X(dL_x(Y))f - Y(dL_x(X))f \\
                    &= (XY - YX)f = [X,Y]f
    \end{aligned}
  \]
\end{proof}

This allows us to endow the tangent space at the identity with the operation $[X_e, Y_e] = [X,Y]_e$ which makes it the lie algebra.

\begin{defn}
  A Riemannian metric on $G$ is left (right) invariant  if
  \[
    \inn*{u}{v}_y = \inn*{dL_x u}{dL_x v}_{xy}
  \]
  for all $x,y \in G$, $u,v \in T_yG$.
\end{defn}

\begin{rem}
  Note that if we have an inner product on the lie algebra $\mathcal{G}$, this induces a left (right) invariant metric on $G$, as
  \[
    \inn*{u}{v}_x = \inn*{dL_{x^{-1}}(u)}{dL_{x^{-1}}(v)}_e
  \]
  for $u,v \in T_xG$, $x \in G$.
\end{rem}

\section*{Adjoint Representation}%

Consider the mapping given by
\[
  \begin{aligned}
    R_{x^{-1}}L_x: &G \rightarrow G \\
                   &y \mapsto xyx^{-1}
  \end{aligned}
\]
This mapping is a diffeomorphism that keeps $e$ fixed also known at the inner automorphism. 
\begin{defn}
The differential at the identity  of the inner automorphism $d(R_{x^{-1}}L_x)_e$  is called the adjoint
\[
  \begin{aligned}
    Ad_x: &\mathcal{G} \rightarrow \mathcal{G} \\
          &v \mapsto d(R_{x^{-1}}L_x)_ev
  \end{aligned}
\]
\end{defn}

We note that the differential is a linear map (transformation) and so this provides a canonical group representation of $G$ by how it tranforms lie algebra $\mathcal{G}$. In this way we consider
\[
  \begin{aligned}
    Ad: &G \rightarrow Aut(\mathcal{G}) \\
        &x \mapsto Ad_x
  \end{aligned}
\]
where $Aut(\mathcal{G})$ is the automorphism group of $\mathcal{G}$ (i.e. invertible linear transformations). 
Note: I don't know any representation theory so I don't know why this is important. 

Now let $X,Y$ be left invariant vector fields. Let $x_t$ be the flow under $X$. Recall that
\[
  [X,Y]_y = \lim_{t \rightarrow 0} \frac{1}{t} \left[Y - dx_t Y \right]_{x_t(y)}
\]
we note that
\[
  x_t(y) = x_t(L_y(e)) = yx_t(e) = R_{x_t(e)}(y)
\]
so $dx_t = dR_{x_t(e)}$ and
\[
  \begin{aligned}
    [X,Y]_y &= \lim_{t \rightarrow 0} \left[Y - dR_{x_t(e)}Y\right] \\
          &= \lim_{t \rightarrow 0} \left[Y - dR_{x_t(e)}dL_{x_t(e)}Y\right] \\
          &= \lim_{t \rightarrow 0} \left[Y - Ad_{x_t^{-1}(e)}Y\right]_{x_t(y)} \\
  \end{aligned}
\]
I'm sure this is cool, but I don't know why.

\end{document}
